---
title: "Luangwa Lion Integrated Population Model"
subtitle: "Survival, reproduction and population dynamics inside and outside South Luangwa National Park"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    embed-resources: true
    fig-width: 8
    fig-height: 5
    code-fold: true
    code-summary: "Show code"
    tbl-cap-location: top
params:
  ipm_tag: "_yrprop"
execute:
  echo: false
  warning: false
  message: false
---

```{r rootdir}
#| include: false
# Run the remaining chunks from the repository root, so the paths below match
# the ones used by the scripts in R/.
#
# This was previously handled by `execute-dir: project` in _quarto.yml, but
# turning the folder into a Quarto project makes Quarto open a Deno KV cache
# under .quarto/, and that fails with "ERROR: disk I/O error" on this drive.
# Setting root.dir here needs no project file and no cache.
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(coda)

`%||%` <- function(a, b) if (is.null(a)) b else a

# Which fit to render.  output/ is gitignored, so it does NOT switch with the
# git branch -- naming the file explicitly keeps this branch's report pointed
# at this branch's results without disturbing main's.  "" is main's fit.
#
# Passed as a Quarto parameter so one source renders every fit:
#   quarto render ipm_report.qmd
#   quarto render ipm_report.qmd -P ipm_tag:_stoch --output ipm_report_stochastic.html
# Keeping a single source means the two reports cannot drift apart; the prose
# that depends on model structure is switched on the flags detected below.
IPM_TAG <- params$ipm_tag %||% "_yrprop"

ipm <- readRDS(sprintf("output/ipm_samples%s.RDS", IPM_TAG))
s <- ipm$summary
sy <- ipm$study_years
Pop <- ipm$Pop
set. <- ipm$settings

dat <- readRDS("data/lion_ipm_data.RDS")
jd <- dat$jags.data
fecd <- dat$jags.data.fec
cc <- read.csv("output/Pop_estimates_monthly_independent.csv")

# Observed movement and interval lengths, recomputed from the data bundle the
# same way R/04_ipm.R does, so the report cannot quote stale hard-coded values
# if the protection rule or the occasion structure changes.
okey <- dat$occasion_key
stratum_by_year <- matrix(NA_integer_, nrow(jd$y), length(sy))
for (t in seq_along(sy)) {
  cols <- okey$occasion[okey$year == sy[t]]
  seen <- which(rowSums(jd$y[, cols, drop = FALSE]) > 0)
  for (i in seen) {
    det <- cols[jd$y[i, cols] == 1]
    stratum_by_year[i, t] <- if (mean(jd$area[i, det] == 1) > 0.5) 1L else 2L
  }
}
mv <- list(n12 = 0L, n1 = 0L, n21 = 0L, n2 = 0L)
for (t in seq_len(length(sy) - 1)) {
  a <- stratum_by_year[, t]
  b <- stratum_by_year[, t + 1]
  ok <- !is.na(a) & !is.na(b)
  mv$n1 <- mv$n1 + sum(a[ok] == 1)
  mv$n2 <- mv$n2 + sum(a[ok] == 2)
  mv$n12 <- mv$n12 + sum(a[ok] == 1 & b[ok] == 2)
  mv$n21 <- mv$n21 + sum(a[ok] == 2 & b[ok] == 1)
}
obs_p12 <- mv$n12 / mv$n1
obs_p21 <- mv$n21 / mv$n2
obs_equil <- obs_p21 / (obs_p12 + obs_p21)
obs_prop_in <- mean(stratum_by_year == 1, na.rm = TRUE)

n_bins_rep <- jd$n.occasions / length(sy)
int_len_rep <- as.numeric(tapply(
  as.numeric(diff(okey$mid_date)) / 365.25,
  ((seq_len(jd$n.occasions - 1) - 1) %% n_bins_rep) + 1,
  mean
))
# Detect whether year effects were on from the samples rather than the settings
# list, so this works for fits saved before the switches were recorded.  When
# use.yr.re is 0 every eps.yr is deterministically zero, so its SD is zero.
year_re_on <- any(s$sd[grepl("eps.yr[", s$parameter, fixed = TRUE)] > 0)
fec_re_on <- any(s$sd[grepl("eps.fec[", s$parameter, fixed = TRUE)] > 0)
proc_on <- any(grepl("sigma.proc", s$parameter, fixed = TRUE)) &&
  any(s$sd[grepl("eps.proc[", s$parameter, fixed = TRUE)] > 0)

# All posterior draws, chains pooled -- needed for the density figures
draws <- as.matrix(ipm$samples)

# pull a scalar parameter by name
gv <- function(p, col = "mean") s[[col]][match(p, s$parameter)]

# rows matching a regex, with index columns parsed out of "name[i, j, k]"
grab <- function(pattern) {
  r <- s[grepl(pattern, s$parameter), , drop = FALSE]
  idx <- regmatches(r$parameter, regexpr("\\[.*\\]", r$parameter))
  idx <- gsub("\\[|\\]", "", idx)
  parts <- strsplit(idx, ",\\s*")
  maxn <- max(lengths(parts))
  for (k in seq_len(maxn)) {
    r[[paste0("i", k)]] <- as.integer(vapply(
      parts,
      function(x) if (length(x) >= k) x[k] else NA_character_,
      ""
    ))
  }
  r
}

age_labs <- c("Cub (0–1.99)", "Subadult (2–3.99)", "Prime (4–5.99)", "Old (6+)")
sex_labs <- c("Female", "Male")
prot_labs <- c("Inside park", "Outside park")

pal <- c("Inside park" = "#16b900", "Outside park" = "#014fa8")

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      legend.position = "top",
      strip.text = element_text(face = "bold")
    )
)

tbl_style <- function(x) {
  # A system font stack, not google_font(): with embed-resources the whole
  # font family is inlined as base64 in every table, which took this report
  # to 17 MB, of which 15.7 MB was fonts.
  x |>
    opt_table_font(font = gt::default_fonts()) |>
    tab_options(
      table.font.size = px(14),
      heading.title.font.size = px(17),
      heading.subtitle.font.size = px(13),
      column_labels.font.weight = "bold",
      data_row.padding = px(5),
      table.border.top.style = "none",
      source_notes.font.size = px(11)
    )
}
```

::: {.callout-note appearance="minimal"}
Generated from `output/ipm_samples`r if (nchar(IPM_TAG)) IPM_TAG else ""``.RDS`.
`r set.$N_CHAINS %||% 3` chains × `r format(set.$N_ITER, big.mark = ",")` iterations,
burn-in `r format(set.$N_BURNIN, big.mark = ",")`, thin `r set.$N_THIN`.
:::

# The model

This IPM follows Creel et al. (2024,
*Conservation Science and Practice*), fitted to
`r format(nrow(jd$y), big.mark = ",")` individually identified lions monitored
in and around South Luangwa National Park from `r min(sy)` to `r max(sy)`.

```{r}
#| label: tbl-components
#| tbl-cap: "The four component models and the data each one uses."
tibble::tibble(
  Component = c(
    "Cormack–Jolly–Seber",
    "Zero-inflated Poisson",
    "Leslie projection",
    if (proc_on) "Gaussian state-space + process error" else "Gaussian state-space"
  ),
  Data = c(
    sprintf("%s capture histories over %d occasions", format(jd$nind, big.mark = ","), jd$n.occasions),
    sprintf("%d adult-female-years of cub counts", fecd$n),
    "—",
    sprintf("%d annual closed-capture estimates", sum(!is.na(Pop)))
  ),
  Estimates = c(
    "Apparent survival by age class, sex and protection",
    "Fecundity by protection level",
    "Age structure, dispersal between strata, abundance through time",
    if (proc_on) {
      "Population size as a latent variable; growth rate; observation error separated from real fluctuation"
    } else {
      "Population size as a latent variable; growth rate"
    }
  )
) |>
  gt() |>
  tab_header(
    title = "Model structure",
    subtitle = "All components share one likelihood"
  ) |>
  cols_width(Component ~ px(190)) |>
  tbl_style()
```

**Occasions.** Four 2-month bins per year (May–Jun, Jul–Aug, Sep–Oct, Nov–Dec)
give `r jd$n.occasions` occasions across `r length(sy)` years. Creel et al.
used three bins covering May–Oct; the Luangwa season runs later, and the fourth
bin raises retention from roughly 75% to 90% of individual-months.

Survival is estimated on the **annual** scale, and the per-interval value the
CJS needs is derived from it. That ordering matters twice over. The intervals
are not equal — the gaps between bin midpoints run
`r paste(sprintf("%.3f", int_len_rep), collapse = ", ")` of a year, the last
spanning the wet season — so treating each as a quarter of a year, as the
published model does, overstates survival across the long gap and understates
it across the three short ones. And a year random effect placed on
per-interval survival is not identifiable, because per-interval survival sits
close to 1 where the likelihood is flat; on the annual scale it is.

**Protection.** Every sighting was overlaid on the WDPA park boundary, and each
lion classified by the majority of **all its sightings in a year**, applied to
every occasion in that year. An occasion-level majority was tried first, but
44.9% of those classifications rested on a single sighting against 25.4% at the
annual level, so one opportunistic excursion outside the boundary could define
a whole 2-month bin and then propagate through carry-forward. Ties are broken
by a coin flip, which makes the assignment mildly seed-dependent.

`r sum(apply(jd$area, 1, function(r) length(unique(r)) > 1))` of `r jd$nind`
lions changed protected status during the study, which is what separates the
protection effect from fixed differences between locations. The change of rule
moved 5.5% of lion-occasions but left every estimate essentially untouched —
`beta.prot` shifted by 0.027, a fifth of its posterior SD.

**Fecundity** is first-year cubs per adult female per year. Litter counts and
sizes are observed; maternal identity was known for only a third of litters, so
litters were allocated within their pride — named mothers first, the remainder
at random among the other adult females, one each.

**Movement.** The projection is not a pair of closed sub-populations. Lions move
between strata constantly, and the model estimates annual transition rates from
lion-years where the stratum was observed in both year *t* and *t+1*, then moves
survivors accordingly. This matters more than it might appear — see
@sec-movement.

```{r}
#| label: tbl-movement
#| tbl-cap: "Movement between strata. The model estimates the transition rates; the observed column is the raw count from the capture histories and is not fitted to directly, so the agreement is a check rather than a tautology."
mv <- s[match(c("psi12", "psi21", "prop.equil"), s$parameter), ]
tibble::tibble(
  Quantity = c(
    "Annual P(move outside | inside)",
    "Annual P(move inside | outside)",
    "Equilibrium proportion inside"
  ),
  Estimate = mv$q50,
  Lower = mv$q2.5,
  Upper = mv$q97.5,
  Observed = c(obs_p12, obs_p21, obs_equil)
) |>
  gt() |>
  tab_header(
    title = "Dispersal between strata",
    subtitle = "Estimated inside the IPM, against the raw capture-history counts"
  ) |>
  fmt_number(columns = c(Estimate, Lower, Upper, Observed), decimals = 3) |>
  cols_merge(columns = c(Lower, Upper), pattern = "{1} – {2}") |>
  cols_label(Lower = "95% CrI") |>
  cols_width(Quantity ~ px(280)) |>
  tab_source_note(
    "Rates apply to all ages and both sexes. Cubs move with their pride rather than dispersing independently, so a shared rate is reasonable for them; treating male and female dispersal as equal is a simplification worth relaxing."
  ) |>
  tbl_style()
```

The equilibrium composition implied by these rates --
`r sprintf("%.3f", gv("prop.equil", "q50"))` inside -- matches the
`r sprintf("%.3f", obs_prop_in)` actually observed. Composition here is set by
dispersal, not by differential demography.

# Headline results

```{r}
#| label: tbl-headline
#| tbl-cap: "Headline estimates. Intervals are 95% credible intervals."
b_sex <- s[s$parameter == "beta.sex", ]
b_prot <- s[s$parameter == "beta.prot", ]
fec1 <- s[s$parameter == "fec[1]", ]
fec2 <- s[s$parameter == "fec[2]", ]
lam <- grab("^lambda\\.tot")
lam_last <- lam[nrow(lam), ]
nt <- grab("^Ntot")

tibble::tibble(
  Quantity = c(
    "Effect of being outside the park on logit(survival)",
    "Effect of being male on logit(survival)",
    "Fecundity, inside park (cubs / adult female / year)",
    "Fecundity, outside park",
    sprintf("Population size, %d", max(sy)),
    sprintf("Population growth rate, %d–%d", max(sy) - 1, max(sy))
  ),
  Estimate = c(
    b_prot$q50, b_sex$q50, fec1$q50, fec2$q50,
    nt$q50[nrow(nt)], lam_last$q50
  ),
  Lower = c(
    b_prot$q2.5, b_sex$q2.5, fec1$q2.5, fec2$q2.5,
    nt$q2.5[nrow(nt)], lam_last$q2.5
  ),
  Upper = c(
    b_prot$q97.5, b_sex$q97.5, fec1$q97.5, fec2$q97.5,
    nt$q97.5[nrow(nt)], lam_last$q97.5
  )
) |>
  mutate(
    excludes_zero = ifelse(
      Quantity %in% c(
        "Effect of being outside the park on logit(survival)",
        "Effect of being male on logit(survival)"
      ),
      sign(Lower) == sign(Upper),
      NA
    )
  ) |>
  gt() |>
  tab_header(title = "Headline estimates") |>
  fmt_number(columns = c(Estimate, Lower, Upper), decimals = 3) |>
  cols_merge(columns = c(Lower, Upper), pattern = "{1} – {2}") |>
  cols_label(Lower = "95% CrI") |>
  cols_hide(excludes_zero) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Estimate, rows = !is.na(excludes_zero) & excludes_zero)
  ) |>
  tab_footnote(
    footnote = "Credible interval excludes zero.",
    locations = cells_body(
      columns = Estimate,
      rows = !is.na(excludes_zero) & excludes_zero
    )
  ) |>
  cols_width(Quantity ~ px(340)) |>
  tbl_style()
```

Lions ranging outside the park have substantially lower apparent survival. Creel et al. found only a weak
tendency in the same direction in Kafue. Protected status shows no effect on fecundity.

::: {.callout-warning appearance="simple"}
Apparent survival is modeled here, which may confound mortality with permanent emigration. Lions ranging
outside the park are plausibly may also be more likely to leave the study area
altogether, so part of the protection effect may be dispersal rather than death?
:::

# Survival

```{r}
#| label: tbl-survival
#| tbl-cap: "Annual apparent survival by age class, sex and protection stratum."
surv <- grab("^surv\\.annual") |>
  transmute(
    age = age_labs[i1],
    sex = sex_labs[i2],
    prot = prot_labs[i3],
    est = q50,
    lo = q2.5,
    hi = q97.5
  )

surv |>
  mutate(cell = sprintf("%.3f (%.3f–%.3f)", est, lo, hi)) |>
  select(age, sex, prot, cell) |>
  pivot_wider(names_from = c(prot, sex), values_from = cell, names_sep = "|") |>
  gt(rowname_col = "age") |>
  tab_header(
    title = "Annual apparent survival",
    subtitle = "Posterior median (95% credible interval)"
  ) |>
  tab_spanner(label = "Inside park", columns = starts_with("Inside")) |>
  tab_spanner(label = "Outside park", columns = starts_with("Outside")) |>
  cols_label(
    `Inside park|Female` = "Female",
    `Inside park|Male` = "Male",
    `Outside park|Female` = "Female",
    `Outside park|Male` = "Male"
  ) |>
  tab_stubhead(label = "Age class") |>
  tab_style(
    style = cell_fill(color = "#eef5ef"),
    locations = cells_body(columns = starts_with("Inside"))
  ) |>
  tab_source_note(
    "The model estimates additive effects of sex and protection on the logit scale, with no interaction."
  ) |>
  tbl_style()
```

```{r}
#| label: fig-survival
#| fig-cap: "Annual apparent survival. Points are posterior medians, bars 95% credible intervals."
#| fig-height: 4.5
surv |>
  mutate(age = factor(age, levels = age_labs)) |>
  ggplot(aes(x = age, y = est, colour = prot, shape = sex)) +
  geom_linerange(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.55),
    linewidth = 0.7
  ) +
  geom_point(position = position_dodge(width = 0.55), size = 2.8) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_shape_manual(values = c(Female = 16, Male = 17), name = NULL) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = NULL, y = "Annual apparent survival") +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

Survival rises from cub to prime adult and falls again in old age, in every
sex-by-protection combination. The gap between strata is constant on the logit
scale by construction: the model fits additive effects with no interaction, as
in the published analysis.

# Reproduction

```{r}
#| label: tbl-fecundity
#| tbl-cap: "Fecundity component. The zero-inflated Poisson separates whether a female bred at all from how many cubs she raised."
fec_rows <- s[s$parameter %in% c("psi", "alpha", "gamma[2]", "fec[1]", "fec[2]"), ]
fec_rows$label <- c(
  alpha = "log mean litter contribution, inside park (α)",
  psi = "Probability a female bred (ψ)",
  `gamma[2]` = "Difference for outside park, log scale (γ₂)",
  `fec[1]` = "Fecundity, inside park",
  `fec[2]` = "Fecundity, outside park"
)[fec_rows$parameter]

fec_rows |>
  select(label, q50, q2.5, q97.5, Rhat) |>
  gt() |>
  tab_header(
    title = "Fecundity",
    subtitle = "First-year cubs per adult female per year"
  ) |>
  fmt_number(columns = c(q50, q2.5, q97.5), decimals = 3) |>
  fmt_number(columns = Rhat, decimals = 3) |>
  cols_merge(columns = c(q2.5, q97.5), pattern = "{1} – {2}") |>
  cols_label(label = "", q50 = "Median", q2.5 = "95% CrI") |>
  cols_width(label ~ px(320)) |>
  tab_source_note(
    sprintf(
      "Based on %d adult-female-years, %d–%d. Maternal identity was imputed within pride-years for the litters where it was unknown.",
      fecd$n, min(dat$settings$FEC_YEARS), max(dat$settings$FEC_YEARS)
    )
  ) |>
  tbl_style()
```

Only about `r round(100 * gv("psi"))`% of adult-female-years involved breeding (the zero inflation doing its job) — a lioness with surviving cubs from
the previous year does not conceive. Among those that did breed, the Poisson
mean was `r round(exp(gv("alpha")), 2)` cubs.

# Posterior distributions

Laid out as in Figure 3 of Creel et al.: fecundity on top, then annual apparent
survival for each age class, with both sexes and both strata overlaid. Showing
the full posterior rather than an interval makes the degree of overlap between
strata visible — which is the quantity of interest when asking whether
protection matters.

```{r}
#| label: fig-posteriors
#| fig-cap: "Posterior distributions by protection level. (a) Adult female fecundity, as first-year cubs per female. (b) Annual apparent survival for the four age classes; within each panel the male distributions sit to the left of the female ones."
#| fig-height: 8
library(patchwork)

fec_draws <- data.frame(
  value = c(draws[, "fec[1]"], draws[, "fec[2]"]),
  prot = rep(prot_labs, each = nrow(draws))
)

p_fec <- ggplot(fec_draws, aes(value, fill = prot, colour = prot)) +
  geom_density(alpha = 0.45, linewidth = 0.5) +
  scale_fill_manual(values = pal, name = NULL) +
  scale_colour_manual(values = pal, name = NULL) +
  labs(
    x = "First-year cubs per adult female per year",
    y = "Posterior density",
    tag = "a"
  )

grid <- expand.grid(a = 1:4, sx = 1:2, pr = 1:2)
surv_draws <- do.call(rbind, lapply(seq_len(nrow(grid)), function(k) {
  a <- grid$a[k]
  sx <- grid$sx[k]
  pr <- grid$pr[k]
  data.frame(
    value = draws[, sprintf("surv.annual[%d,%d,%d]", a, sx, pr)],
    age = age_labs[a],
    sex = sex_labs[sx],
    prot = prot_labs[pr]
  )
}))
surv_draws$age <- factor(surv_draws$age, levels = age_labs)

p_surv <- ggplot(
  surv_draws,
  aes(value, fill = prot, colour = prot, linetype = sex)
) +
  geom_density(alpha = 0.3, linewidth = 0.5) +
  facet_wrap(~age, nrow = 2) +
  scale_fill_manual(values = pal, name = NULL) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_linetype_manual(
    values = c(Female = "solid", Male = "22"),
    name = NULL
  ) +
  labs(x = "Annual apparent survival", y = "Posterior density", tag = "b")

p_fec / p_surv + plot_layout(heights = c(1, 2))
```

The separation between strata is consistent across all four age classes,
because the model fits an additive protection effect on the logit scale with no
age interaction — the panels show how that constant shift translates into
different absolute gaps at different baseline survival rates. Fecundity, by
contrast, shows two distributions that overlap almost entirely.

# Between-year variation

Survival and fecundity each carry a year random effect, on the annual logit and
log scales respectively. Both are shrunk toward zero by the data rather than
imposed.

```{r}
#| label: fig-yeareffects
#| fig-cap: "Year random effect on annual survival, on the logit scale. Points are posterior medians, bars 95% credible intervals; a year whose interval clears zero is filled."
#| fig-height: 4
ey <- s[grepl("eps.yr[", s$parameter, fixed = TRUE), ]
ey <- ey[order(as.integer(gsub("[^0-9]", "", ey$parameter))), ]
ey_df <- data.frame(
  year = sy,
  est = ey$q50,
  lo = ey$q2.5,
  hi = ey$q97.5
)
ey_df$clear <- sign(ey_df$lo) == sign(ey_df$hi)

ggplot(ey_df, aes(year, est)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  geom_linerange(aes(ymin = lo, ymax = hi), colour = "grey45") +
  geom_point(aes(fill = clear), shape = 21, size = 2.6, colour = "grey20") +
  scale_fill_manual(values = c(`TRUE` = "#b2182b", `FALSE` = "white"),
                    guide = "none") +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  labs(x = NULL, y = "Year effect on logit(annual survival)")
```

```{r}
#| label: tbl-varcomp
#| tbl-cap: "Variance components. sigma.yr and sigma.fec.yr are the between-year SDs of survival and fecundity on their respective link scales."
vc <- s[s$parameter %in% c("sigma.yr", "sigma.fec.yr", "sigma.p", "sigma.c"), ]
vc$label <- c(
  sigma.yr = "Between-year SD, survival (logit)",
  sigma.fec.yr = "Between-year SD, fecundity (log)",
  sigma.p = "Between-individual SD, detection (logit)",
  sigma.c = "Sampling SD, population estimates"
)[vc$parameter]
vc |>
  select(label, q50, q2.5, q97.5, Rhat, n.eff) |>
  gt() |>
  tab_header(title = "Variance components") |>
  fmt_number(columns = c(q50, q2.5, q97.5), decimals = 3) |>
  fmt_number(columns = Rhat, decimals = 3) |>
  fmt_number(columns = n.eff, decimals = 0) |>
  cols_merge(columns = c(q2.5, q97.5), pattern = "{1} – {2}") |>
  cols_label(label = "", q50 = "Median", q2.5 = "95% CrI") |>
  cols_width(label ~ px(300)) |>
  tbl_style()
```

`sigma.yr` is `r sprintf("%.3f", gv("sigma.yr", "q50"))`
(`r sprintf("%.3f", gv("sigma.yr", "q2.5"))` –
`r sprintf("%.3f", gv("sigma.yr", "q97.5"))`). For a prime adult female inside
the park, mean annual survival of
`r sprintf("%.3f", gv("surv.annual[3,1,1]", "q50"))` and a ±1 SD year effect
span roughly
`r sprintf("%.2f", plogis(qlogis(gv("surv.annual[3,1,1]", "q50")) - gv("sigma.yr", "q50")))`
to
`r sprintf("%.2f", plogis(qlogis(gv("surv.annual[3,1,1]", "q50")) + gv("sigma.yr", "q50")))`.
Only `r sy[which(ey_df$clear)]` has a year effect whose interval clears zero.

::: {.callout-important appearance="simple"}
**This effect is only estimable on the annual scale.** Placed on per-interval
survival — the natural way to write it, since that is what the CJS models —
`sigma.yr` is not identified at all. Per-interval survival is about 0.96, which
on the logit scale sits close enough to the boundary that large changes barely
move the probability; the likelihood is flat there and `sigma.yr` runs to
whatever bound the prior supplies. Widening the prior from `dunif(0,5)` to
`dunif(0,20)` moved the estimate from 3.9 to 15.9, which is how this was
diagnosed. On the annual scale, where survival is around 0.85, it is well
identified and R̂ = `r sprintf("%.3f", gv("sigma.yr", "Rhat"))`.
:::

`sigma.fec.yr` = `r sprintf("%.3f", gv("sigma.fec.yr", "q50"))`
(`r sprintf("%.3f", gv("sigma.fec.yr", "q2.5"))` –
`r sprintf("%.3f", gv("sigma.fec.yr", "q97.5"))`) is shrunk hard toward zero.
With `r fecd$n` female-years across `r length(unique(dat$fec_data$year))` years
and `r sprintf("%.0f%%", 100 * mean(fecd$C == 0))` of them zeros, between-year
variation cannot be separated from the zero-inflation parameter ψ. Read that as
*not detectable in these data*, not as evidence that fecundity is constant.

# Population size and growth

```{r}
#| label: fig-popsize
#| fig-cap: "Population size from the IPM against the closed-capture estimates it was fitted to. Open points are years excluded by the survey-effort restriction and contribute no likelihood."
nt_df <- tibble::tibble(
  year = sy,
  est = nt$q50,
  lo = nt$q2.5,
  hi = nt$q97.5,
  Pop = Pop
)
cc_df <- cc |> transmute(year, N = N_median, used = effort_ok)

ggplot(nt_df, aes(year)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.45) +
  geom_line(aes(y = est), linewidth = 0.9) +
  geom_point(
    data = cc_df,
    aes(y = N, shape = used, fill = used),
    size = 2.6,
    colour = "#b2182b"
  ) +
  scale_shape_manual(
    values = c(`TRUE` = 21, `FALSE` = 21),
    labels = c("excluded (low effort)", "used"),
    name = NULL
  ) +
  scale_fill_manual(
    values = c(`TRUE` = "#b2182b", `FALSE` = "white"),
    labels = c("excluded (low effort)", "used"),
    name = NULL
  ) +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  labs(x = NULL, y = "Lions")
```

```{r}
#| label: tbl-popsize
#| tbl-cap: "Population size by year, with the closed-capture estimate supplied to the state-space model."
nt_df |>
  left_join(cc_df, by = "year") |>
  transmute(
    year,
    N,
    used,
    est,
    lo,
    hi,
    dens = est / (set.$AREA_KM2 %||% NA_real_) * 100
  ) |>
  gt() |>
  tab_header(
    title = "Population size",
    subtitle = "IPM posterior against the closed-capture input"
  ) |>
  fmt_number(columns = c(N, est, lo, hi), decimals = 0) |>
  fmt_number(columns = dens, decimals = 2) |>
  sub_missing(missing_text = "—") |>
  cols_merge(columns = c(lo, hi), pattern = "{1} – {2}") |>
  cols_label(
    year = "Year",
    N = "Closed capture",
    used = "Used",
    est = "IPM median",
    lo = "95% CrI",
    dens = "per 100 km²"
  ) |>
  text_transform(
    locations = cells_body(columns = used),
    fn = function(x) ifelse(x == "TRUE", "✓", "")
  ) |>
  tab_style(
    style = cell_text(color = "#999999"),
    locations = cells_body(rows = !used)
  ) |>
  tab_source_note(
    sprintf(
      "Density uses a placeholder area of %s km². Replace AREA_KM2 in R/04_ipm.R with the kernel-utilisation estimate before reporting density.",
      format(set.$AREA_KM2 %||% NA, big.mark = ",")
    )
  ) |>
  tbl_style()
```

```{r}
#| label: fig-lambda
#| fig-cap: "Annual population growth rate. Survival carries a year random effect that feeds through into the projection, so this varies with the estimated year-to-year survival rather than being constant. Fecundity is held at its marginal mean -- see the note below."
#| fig-height: 4
lam_df <- tibble::tibble(
  year = sy[-length(sy)],
  est = lam$q50,
  lo = lam$q2.5,
  hi = lam$q97.5
)
ggplot(lam_df, aes(year, est)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.45) +
  geom_line(linewidth = 0.9) +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  labs(x = NULL, y = expression(lambda))
```

::: {.callout-important appearance="simple"}
**What this variation does and does not include.** $\lambda$ varies here because
the year random effect on survival feeds through into the projection. It tracks
that effect closely — the correlation between $\lambda$ and `eps.yr` is 0.99 —
so this is year-to-year variation in growth attributable to *measured survival*.

```{r}
#| output: asis
if (proc_on) {
  cat(
    "The projection also carries process error, so variation beyond what",
    "survival explains has somewhere to go rather than being charged to",
    "observation error. What remains outside the model is attribution:",
    "`sigma.proc` is a flat multiplier on abundance and cannot say which vital",
    "rate moved. Fecundity is still held at its marginal mean across years,",
    "because the litter records stop in 2015 and a year-specific value for",
    "2016-2023 would come from the prior rather than from data, so recruitment",
    "variation lands in `sigma.proc` undifferentiated.\n"
  )
} else {
  cat(
    "It is not total environmental stochasticity. Fecundity is held at its",
    "marginal mean across years, because the litter records stop in 2015 and a",
    "year-specific value for 2016-2023 would come from the prior rather than",
    "from data. And the projection is deterministic given the vital rates:",
    "there is no process error, so every departure of the real population from",
    "the fitted trajectory is charged to observation error in `sigma.c`.",
    "Variation in recruitment -- often the larger component for lions -- has",
    "nowhere to go in this model.\n"
  )
}
```
:::

```{r}
#| output: asis
#| label: sec-variance
# Everything here is emitted from inside the conditional, including the table.
# A separate chunk with `eval: !expr proc_on` does not work: knitr skips the
# code but Quarto still emits the caption, leaving an orphaned "Observation
# error against process error" heading above nothing in the non-stochastic
# report.
if (proc_on) {
  n_mean <- mean(s$q50[grepl("Ntot[", s$parameter, fixed = TRUE)])
  sc <- gv("sigma.c", "q50")
  sp <- gv("sigma.proc", "q50")

  cat("\n## Splitting observation error from real fluctuation\n\n")
  cat(
    "Without process error every departure of the population from the fitted",
    "trajectory is charged to `sigma.c`, which conflates two different things:",
    "how imprecise the closed-capture estimates are, and how much the",
    "population actually fluctuates. Fitting both separates them.\n\n"
  )

  tab <- tibble::tibble(
    Source = c(
      "Observation error in the closed-capture estimates",
      "Process error: real year-to-year fluctuation"
    ),
    Parameter = c("sigma.c", "sigma.proc"),
    Estimate = c(sc, sp),
    Lower = c(gv("sigma.c", "q2.5"), gv("sigma.proc", "q2.5")),
    Upper = c(gv("sigma.c", "q97.5"), gv("sigma.proc", "q97.5")),
    `As lions` = c(sc, n_mean * (exp(sp) - 1))
  ) |>
    gt() |>
    tab_header(
      title = "Variance decomposition",
      subtitle = sprintf("At a mean population of %.0f lions", n_mean)
    ) |>
    fmt_number(columns = c(Estimate, Lower, Upper), decimals = 3) |>
    fmt_number(columns = `As lions`, decimals = 1) |>
    cols_merge(columns = c(Lower, Upper), pattern = "{1} – {2}") |>
    cols_label(Lower = "95% CrI") |>
    cols_width(Source ~ px(300)) |>
    tab_source_note(paste(
      "sigma.c is already in lions; sigma.proc is an SD on the log scale,",
      "shown as the equivalent proportional swing at the mean population size."
    )) |>
    tbl_style()
  cat(gt::as_raw_html(tab), "\n\n")

  cat(sprintf(
    paste(
      "The two are comparable in size -- roughly %.0f lions of measurement",
      "imprecision against %.0f lions of genuine fluctuation, a %.1f%% annual",
      "CV. Fitted without process error the same model puts `sigma.c` at 25.0,",
      "so about a third of what was being called observation error is really",
      "the population moving.\n\n",
      "Two cautions. The lower bound on `sigma.proc` is %.3f, so the data do",
      "not exclude near-zero process variance. And `sigma.c` spans %.1f to",
      "%.1f: with %d population estimates the split between the two is",
      "informed but not sharply determined.\n"
    ),
    sc, n_mean * (exp(sp) - 1), 100 * (exp(sp) - 1),
    gv("sigma.proc", "q2.5"), gv("sigma.c", "q2.5"), gv("sigma.c", "q97.5"),
    sum(!is.na(Pop))
  ))
}
```


Laid out as in Figure 4 of Creel et al.: the growth-rate posterior, then the
density trend with credible intervals. Panel (a) covers the population as a
whole; @sec-movement explains why that is the right unit here.

```{r}
#| label: fig-growth-density
#| fig-cap: "(a) Posterior distribution for the annual population growth rate. A single curve rather than one per protection level: with dispersal in the projection the two strata share a growth rate — see -movement. (b) Density through time; points are posterior means and bars 95% credible intervals."
#| fig-height: 8
last_t <- length(sy) - 1

lam_draws <- data.frame(
  value = draws[, sprintf("lambda.tot[%d]", last_t)],
  grp = "Whole population"
)
pal_lam <- c("Whole population" = "#333333")

p_lam <- ggplot(lam_draws, aes(value, fill = grp, colour = grp)) +
  geom_density(alpha = 0.4, linewidth = 0.5) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey40") +
  scale_fill_manual(values = pal_lam, name = NULL) +
  scale_colour_manual(values = pal_lam, name = NULL) +
  labs(x = expression(lambda), y = "Posterior density", tag = "a")

dens <- grab("^Density")
dens_df <- data.frame(
  year = sy,
  mean = dens$mean * 100,
  lo = dens$q2.5 * 100,
  hi = dens$q97.5 * 100
)

p_dens <- ggplot(dens_df, aes(year, mean)) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    width = 0.25,
    colour = "grey45",
    linewidth = 0.5
  ) +
  geom_point(size = 2.4) +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  labs(
    x = NULL,
    y = expression("Lions per 100 km"^2),
    tag = "b"
  )

p_lam / p_dens
```

## Why one growth rate and not two {#sec-movement}

Creel et al.'s Figure 4a shows λ separately for high and low protection. This
model reports a single λ — not because the split could not be estimated, but
because the model with dispersal shows there is no split to report.

```{r}
#| label: tbl-lambda-strata
#| tbl-cap: "Stratum-specific growth rates in the final year. These are separate parameters in the model, estimated independently, and they converge on the same value."
lp <- s[s$parameter %in% c(
  sprintf("lambda.prot[1,%d]", length(sy) - 1),
  sprintf("lambda.prot[2,%d]", length(sy) - 1),
  sprintf("lambda.tot[%d]", length(sy) - 1)
), ]
lp$label <- c(
  "Inside park", "Outside park", "Whole population"
)[match(
  sub("\\[.*", "", lp$parameter),
  c("lambda.prot", "lambda.prot", "lambda.tot")
) + c(0, 1, 0)[seq_len(nrow(lp))]]

data.frame(
  Stratum = c("Inside park", "Outside park", "Whole population"),
  lambda = c(
    gv(sprintf("lambda.prot[1,%d]", length(sy) - 1), "q50"),
    gv(sprintf("lambda.prot[2,%d]", length(sy) - 1), "q50"),
    gv(sprintf("lambda.tot[%d]", length(sy) - 1), "q50")
  ),
  Lower = c(
    gv(sprintf("lambda.prot[1,%d]", length(sy) - 1), "q2.5"),
    gv(sprintf("lambda.prot[2,%d]", length(sy) - 1), "q2.5"),
    gv(sprintf("lambda.tot[%d]", length(sy) - 1), "q2.5")
  ),
  Upper = c(
    gv(sprintf("lambda.prot[1,%d]", length(sy) - 1), "q97.5"),
    gv(sprintf("lambda.prot[2,%d]", length(sy) - 1), "q97.5"),
    gv(sprintf("lambda.tot[%d]", length(sy) - 1), "q97.5")
  ),
  Rhat = c(
    gv(sprintf("lambda.prot[1,%d]", length(sy) - 1), "Rhat"),
    gv(sprintf("lambda.prot[2,%d]", length(sy) - 1), "Rhat"),
    gv(sprintf("lambda.tot[%d]", length(sy) - 1), "Rhat")
  )
) |>
  gt() |>
  tab_header(title = "Growth rate by stratum") |>
  fmt_number(columns = c(lambda, Lower, Upper), decimals = 4) |>
  fmt_number(columns = Rhat, decimals = 3) |>
  cols_merge(columns = c(Lower, Upper), pattern = "{1} – {2}") |>
  cols_label(lambda = "λ", Lower = "95% CrI") |>
  tbl_style()
```

```{r}
#| label: tbl-lambda-diff
#| tbl-cap: "Difference in growth rate between strata, year by year. Taken from the posterior of the difference rather than by comparing two marginal intervals, which would overstate the separation because the two are strongly correlated."
dm <- do.call(rbind, lapply(ipm$samples, as.matrix))
nt_l <- length(sy) - 1
dif <- vapply(seq_len(nt_l), function(t) {
  dm[, sprintf("lambda.prot[1,%d]", t)] - dm[, sprintf("lambda.prot[2,%d]", t)]
}, numeric(nrow(dm)))
qd <- apply(dif, 2, quantile, c(0.025, 0.5, 0.975))
lam_diff <- data.frame(
  interval = paste0(head(sy, -1), "–", tail(sy, -1)),
  inside = apply(dm[, sprintf("lambda.prot[1,%d]", seq_len(nt_l))], 2, median),
  outside = apply(dm[, sprintf("lambda.prot[2,%d]", seq_len(nt_l))], 2, median),
  diff = qd[2, ], lo = qd[1, ], hi = qd[3, ]
)
lam_diff$clear <- sign(lam_diff$lo) == sign(lam_diff$hi)

lam_diff |>
  gt() |>
  tab_header(title = "Growth rate, inside vs outside the park") |>
  fmt_number(columns = c(inside, outside, diff, lo, hi), decimals = 4) |>
  cols_merge(columns = c(lo, hi), pattern = "{1} – {2}") |>
  cols_label(interval = "", inside = "Inside", outside = "Outside",
             diff = "Difference", lo = "95% CrI") |>
  cols_hide(clear) |>
  tab_source_note(sprintf(
    "The two series correlate %.4f. The difference excludes zero in %d of %d intervals.",
    cor(lam_diff$inside, lam_diff$outside), sum(lam_diff$clear), nt_l
  )) |>
  tbl_style()
```

The two strata track each other almost exactly: the series correlate
`r sprintf("%.4f", cor(lam_diff$inside, lam_diff$outside))`, the largest gap in
any year is `r sprintf("%.3f", max(abs(lam_diff$inside - lam_diff$outside)))`,
and the difference excludes zero in
**`r sum(lam_diff$clear)` of `r nt_l` intervals**.

That is close to a structural necessity rather than a coincidence. Dispersal
holds the composition near its equilibrium, and a population whose composition
is stationary must grow at one rate throughout. Year-varying survival perturbs
that slightly — under constant vital rates the two were identical to four
decimal places — but never enough to separate them.

"Growth rate of the protected sub-population" is only a distinct quantity if the
sub-populations are demographically closed, and with
`r sum(apply(jd$area, 1, function(r) length(unique(r)) > 1))` of `r jd$nind`
lions changing stratum, these are not.

The protection effect is real and is reported where it acts — on survival, and
to a much weaker degree on fecundity. It does not translate into two separate
population trajectories, because the lions mix.

::: {.callout-note appearance="simple"}
**How this was established.** Fitted without dispersal, the stratum-specific
quantities were unidentified: `Pop[t]` observes only the total, and `Nprot` and
`lambda.prot` returned $\hat{R}$ up to 1.26 with effective sample sizes near 12.
Adding composition data alone made things worse — with no dispersal available
the model matched the observed composition by distorting the vital rates
instead, pushing `beta.prot` to −1.1 and `sigma.c` to ~70.

Adding dispersal fixed both. Vital rates are unchanged from the no-composition
fit (`beta.prot` −0.542 → −0.518, `sigma.c` 29.9 → 29.3), and non-converged
parameters went from 47 of 156 to **0 of 175**. The equality of the two λ values
is therefore a converged result, not a symptom of poor mixing.
:::

::: {.callout-note appearance="simple"}
Density in panel (b) uses the placeholder area of
`r format(set.$AREA_KM2 %||% NA, big.mark = ",")` km². Only the vertical scale
depends on it; the shape of the trend and the relative widths of the intervals
do not.
:::

# Data behind the model

```{r}
#| label: tbl-closedcapture
#| tbl-cap: "Closed-capture model by year. Years were excluded where survey effort was too low, the chain did not converge, or the posterior pressed against the data-augmentation ceiling."
cc |>
  transmute(
    year, n_known, det_per_lion,
    N_median, N_lower, N_upper,
    mean_p, sigma, Rhat_N,
    used = effort_ok
  ) |>
  gt() |>
  tab_header(
    title = "Annual closed-capture estimates",
    subtitle = "Model Mh with data augmentation, monthly occasions"
  ) |>
  fmt_number(columns = c(N_median, N_lower, N_upper), decimals = 0) |>
  fmt_number(columns = c(det_per_lion, mean_p, sigma, Rhat_N), decimals = 2) |>
  cols_merge(columns = c(N_lower, N_upper), pattern = "{1} – {2}") |>
  cols_label(
    year = "Year",
    n_known = "Identified",
    det_per_lion = "Detections / lion",
    N_median = "N",
    N_lower = "95% CrI",
    mean_p = "p",
    sigma = "σ",
    Rhat_N = "R̂",
    used = "Used"
  ) |>
  text_transform(
    locations = cells_body(columns = used),
    fn = function(x) ifelse(x == "TRUE", "✓", "")
  ) |>
  data_color(
    columns = det_per_lion,
    palette = c("#f6e8c3", "#c7eae5", "#5ab4ac")
  ) |>
  tab_style(
    style = cell_text(color = "#999999"),
    locations = cells_body(rows = !used)
  ) |>
  tbl_style()
```

Detections per identified lion is the effort measure the restriction is based
on. It runs from `r sprintf("%.1f", min(cc$det_per_lion))` to
`r sprintf("%.1f", max(cc$det_per_lion))`, and correlates
`r sprintf("%.2f", cor(cc$det_per_lion, cc$N_median / cc$n_known))` with how far
the closed-capture estimate exceeds the number of lions actually identified. In
other words, thinly surveyed years produce inflated abundance estimates, because
an Mh model explains "each lion seen fewer times" by positing more lions it
never saw. With no independent effort covariate available, the only honest
response was to drop those years.

# Convergence

```{r}
#| label: tbl-convergence
#| tbl-cap: "Convergence summary across all monitored parameters."
stoch <- s |> filter(!is.na(sd), sd > 0, !is.na(Rhat))
tibble::tibble(
  Diagnostic = c(
    "Monitored parameters",
    "Maximum R̂",
    "Parameters with R̂ > 1.1",
    "Minimum effective sample size",
    "Median effective sample size"
  ),
  Value = c(
    as.character(nrow(stoch)),
    sprintf("%.4f", max(stoch$Rhat)),
    as.character(sum(stoch$Rhat > 1.1)),
    format(round(min(stoch$n.eff)), big.mark = ","),
    format(round(median(stoch$n.eff)), big.mark = ",")
  )
) |>
  gt() |>
  tab_header(title = "MCMC diagnostics") |>
  cols_align(align = "left", columns = Diagnostic) |>
  tab_style(
    style = cell_text(
      color = ifelse(max(stoch$Rhat) <= 1.1, "#1b7837", "#b2182b"),
      weight = "bold"
    ),
    locations = cells_body(columns = Value, rows = 2:3)
  ) |>
  tbl_style()
```

# Caveats

1. **Apparent survival is not true survival.** Permanent emigration is
   indistinguishable from death in a CJS model, and it is plausibly more common
   outside the park, where the protection effect is measured.
2. **Cub survival is high** relative to published lion estimates because cubs
   enter the data only once individually identified, so mortality in the first
   months of life is invisible.
3. **Fecundity rests on imputed maternity** and covers
   `r min(dat$settings$FEC_YEARS)`–`r max(dat$settings$FEC_YEARS)`, while
   survival covers `r min(sy)`–`r max(sy)`. It is estimated as one rate per
   stratum rather than per year, so the mismatch is survivable, but this is the
   weakest input in the model, and the absent protection effect on fecundity may
   reflect the imputation as much as the biology.
4. **Population size rests on `r sum(!is.na(Pop))` of `r length(sy)` years.**
5. **Density is provisional** until `AREA_KM2` is replaced with a real estimate
   of the area occupied.
6. **This model departs from the published Kafue code in five documented ways**,
   three of which change $\lambda$. See the header of `R/04_ipm.R`. Estimates
   here are not directly comparable with the published figures.
