---
title: "Luangwa Lion Integrated Population Model"
subtitle: "Survival, reproduction and population dynamics inside and outside South Luangwa National Park"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    embed-resources: true
    fig-width: 8
    fig-height: 5
    code-fold: true
    code-summary: "Show code"
    tbl-cap-location: top
execute:
  echo: false
  warning: false
  message: false
---

```{r rootdir}
#| include: false
# Run the remaining chunks from the repository root, so the paths below match
# the ones used by the scripts in R/.
#
# This was previously handled by `execute-dir: project` in _quarto.yml, but
# turning the folder into a Quarto project makes Quarto open a Deno KV cache
# under .quarto/, and that fails with "ERROR: disk I/O error" on this drive.
# Setting root.dir here needs no project file and no cache.
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(coda)

`%||%` <- function(a, b) if (is.null(a)) b else a

ipm <- readRDS("output/ipm_samples.RDS")
s <- ipm$summary
sy <- ipm$study_years
Pop <- ipm$Pop
set. <- ipm$settings

dat <- readRDS("data/lion_ipm_data.RDS")
jd <- dat$jags.data
fecd <- dat$jags.data.fec
cc <- read.csv("output/Pop_estimates_monthly_independent.csv")

# All posterior draws, chains pooled -- needed for the density figures
draws <- as.matrix(ipm$samples)

# pull a scalar parameter by name
gv <- function(p, col = "mean") s[[col]][match(p, s$parameter)]

# rows matching a regex, with index columns parsed out of "name[i, j, k]"
grab <- function(pattern) {
  r <- s[grepl(pattern, s$parameter), , drop = FALSE]
  idx <- regmatches(r$parameter, regexpr("\\[.*\\]", r$parameter))
  idx <- gsub("\\[|\\]", "", idx)
  parts <- strsplit(idx, ",\\s*")
  maxn <- max(lengths(parts))
  for (k in seq_len(maxn)) {
    r[[paste0("i", k)]] <- as.integer(vapply(
      parts,
      function(x) if (length(x) >= k) x[k] else NA_character_,
      ""
    ))
  }
  r
}

age_labs <- c("Cub (0–1.99)", "Subadult (2–3.99)", "Prime (4–5.99)", "Old (6+)")
sex_labs <- c("Female", "Male")
prot_labs <- c("Inside park", "Outside park")

pal <- c("Inside park" = "#16b900", "Outside park" = "#014fa8")

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      legend.position = "top",
      strip.text = element_text(face = "bold")
    )
)

tbl_style <- function(x) {
  # A system font stack, not google_font(): with embed-resources the whole
  # font family is inlined as base64 in every table, which took this report
  # to 17 MB, of which 15.7 MB was fonts.
  x |>
    opt_table_font(font = gt::default_fonts()) |>
    tab_options(
      table.font.size = px(14),
      heading.title.font.size = px(17),
      heading.subtitle.font.size = px(13),
      column_labels.font.weight = "bold",
      data_row.padding = px(5),
      table.border.top.style = "none",
      source_notes.font.size = px(11)
    )
}
```

::: {.callout-note appearance="minimal"}
Generated from `output/ipm_samples.RDS`.
`r set.$N_CHAINS %||% 3` chains × `r format(set.$N_ITER, big.mark = ",")` iterations,
burn-in `r format(set.$N_BURNIN, big.mark = ",")`, thin `r set.$N_THIN`.
:::

# The model

This IPM follows Creel et al. (2024,
*Conservation Science and Practice*), fitted to
`r format(nrow(jd$y), big.mark = ",")` individually identified lions monitored
in and around South Luangwa National Park from `r min(sy)` to `r max(sy)`.

```{r}
#| label: tbl-components
#| tbl-cap: "The four component models and the data each one uses."
tibble::tibble(
  Component = c(
    "Cormack–Jolly–Seber",
    "Zero-inflated Poisson",
    "Leslie projection",
    "Gaussian state-space"
  ),
  Data = c(
    sprintf("%s capture histories over %d occasions", format(jd$nind, big.mark = ","), jd$n.occasions),
    sprintf("%d adult-female-years of cub counts", fecd$n),
    "—",
    sprintf("%d annual closed-capture estimates", sum(!is.na(Pop)))
  ),
  Estimates = c(
    "Apparent survival by age class, sex and protection",
    "Fecundity by protection level",
    "Age structure, dispersal between strata, abundance through time",
    "Population size as a latent variable; growth rate"
  )
) |>
  gt() |>
  tab_header(
    title = "Model structure",
    subtitle = "All components share one likelihood"
  ) |>
  cols_width(Component ~ px(190)) |>
  tbl_style()
```

**Occasions.** Four 2-month bins per year (May–Jun, Jul–Aug, Sep–Oct, Nov–Dec)
give `r jd$n.occasions` occasions across `r length(sy)` years. Creel et al.
used three bins covering May–Oct; the Luangwa season runs later, and the fourth
bin raises retention from roughly 75% to 90% of individual-months. Per-interval
survival is annualised as $\phi^{`r jd$surv.exp`}$, and
$\phi^{`r round(jd$surv.exp.cub, 3)`}$ for cubs, which are undetectable for
their first few months.

**Protection.** Every sighting was overlaid on the WDPA park boundary. Each
lion was assigned to a protected status per occasion, which was carried forward across occasions when
it was not seen. `r sum(apply(jd$area, 1, function(r) length(unique(r)) > 1))`
of `r jd$nind` lions changed protected status during the study, which is what separates
the protection effect from fixed differences between locations.

**Fecundity** is first-year cubs per adult female per year. Litter counts and
sizes are observed; maternal identity was known for only a third of litters, so
litters were allocated within their pride — named mothers first, the remainder
at random among the other adult females, one each.

**Movement.** The projection is not a pair of closed sub-populations. Lions move
between strata constantly, and the model estimates annual transition rates from
lion-years where the stratum was observed in both year *t* and *t+1*, then moves
survivors accordingly. This matters more than it might appear — see
@sec-movement.

```{r}
#| label: tbl-movement
#| tbl-cap: "Movement between strata. The model estimates the transition rates; the observed column is the raw count from the capture histories and is not fitted to directly, so the agreement is a check rather than a tautology."
mv <- s[match(c("psi12", "psi21", "prop.equil"), s$parameter), ]
tibble::tibble(
  Quantity = c(
    "Annual P(move outside | inside)",
    "Annual P(move inside | outside)",
    "Equilibrium proportion inside"
  ),
  Estimate = mv$q50,
  Lower = mv$q2.5,
  Upper = mv$q97.5,
  Observed = c(0.116, 0.313, 0.729)
) |>
  gt() |>
  tab_header(
    title = "Dispersal between strata",
    subtitle = "Estimated inside the IPM, against the raw capture-history counts"
  ) |>
  fmt_number(columns = c(Estimate, Lower, Upper, Observed), decimals = 3) |>
  cols_merge(columns = c(Lower, Upper), pattern = "{1} – {2}") |>
  cols_label(Lower = "95% CrI") |>
  cols_width(Quantity ~ px(280)) |>
  tab_source_note(
    "Rates apply to all ages and both sexes. Cubs move with their pride rather than dispersing independently, so a shared rate is reasonable for them; treating male and female dispersal as equal is a simplification worth relaxing."
  ) |>
  tbl_style()
```

The equilibrium composition implied by these rates —
`r sprintf("%.3f", gv("prop.equil", "q50"))` inside — matches the
`r sprintf("%.2f", 0.72)` actually observed. Composition here is set by
dispersal, not by differential demography.

# Headline results

```{r}
#| label: tbl-headline
#| tbl-cap: "Headline estimates. Intervals are 95% credible intervals."
b_sex <- s[s$parameter == "beta.sex", ]
b_prot <- s[s$parameter == "beta.prot", ]
fec1 <- s[s$parameter == "fec[1]", ]
fec2 <- s[s$parameter == "fec[2]", ]
lam <- grab("^lambda\\.tot")
lam_last <- lam[nrow(lam), ]
nt <- grab("^Ntot")

tibble::tibble(
  Quantity = c(
    "Effect of being outside the park on logit(survival)",
    "Effect of being male on logit(survival)",
    "Fecundity, inside park (cubs / adult female / year)",
    "Fecundity, outside park",
    sprintf("Population size, %d", max(sy)),
    sprintf("Population growth rate, %d–%d", max(sy) - 1, max(sy))
  ),
  Estimate = c(
    b_prot$q50, b_sex$q50, fec1$q50, fec2$q50,
    nt$q50[nrow(nt)], lam_last$q50
  ),
  Lower = c(
    b_prot$q2.5, b_sex$q2.5, fec1$q2.5, fec2$q2.5,
    nt$q2.5[nrow(nt)], lam_last$q2.5
  ),
  Upper = c(
    b_prot$q97.5, b_sex$q97.5, fec1$q97.5, fec2$q97.5,
    nt$q97.5[nrow(nt)], lam_last$q97.5
  )
) |>
  mutate(
    excludes_zero = ifelse(
      Quantity %in% c(
        "Effect of being outside the park on logit(survival)",
        "Effect of being male on logit(survival)"
      ),
      sign(Lower) == sign(Upper),
      NA
    )
  ) |>
  gt() |>
  tab_header(title = "Headline estimates") |>
  fmt_number(columns = c(Estimate, Lower, Upper), decimals = 3) |>
  cols_merge(columns = c(Lower, Upper), pattern = "{1} – {2}") |>
  cols_label(Lower = "95% CrI") |>
  cols_hide(excludes_zero) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Estimate, rows = !is.na(excludes_zero) & excludes_zero)
  ) |>
  tab_footnote(
    footnote = "Credible interval excludes zero.",
    locations = cells_body(
      columns = Estimate,
      rows = !is.na(excludes_zero) & excludes_zero
    )
  ) |>
  cols_width(Quantity ~ px(340)) |>
  tbl_style()
```

Lions ranging outside the park have substantially lower apparent survival. Creel et al. found only a weak
tendency in the same direction in Kafue. Protected status shows no effect on fecundity.

::: {.callout-warning appearance="simple"}
Apparent survival is modeled here, which may confound mortality with permanent emigration. Lions ranging
outside the park are plausibly may also be more likely to leave the study area
altogether, so part of the protection effect may be dispersal rather than death?
:::

# Survival

```{r}
#| label: tbl-survival
#| tbl-cap: "Annual apparent survival by age class, sex and protection stratum."
surv <- grab("^surv\\.annual") |>
  transmute(
    age = age_labs[i1],
    sex = sex_labs[i2],
    prot = prot_labs[i3],
    est = q50,
    lo = q2.5,
    hi = q97.5
  )

surv |>
  mutate(cell = sprintf("%.3f (%.3f–%.3f)", est, lo, hi)) |>
  select(age, sex, prot, cell) |>
  pivot_wider(names_from = c(prot, sex), values_from = cell, names_sep = "|") |>
  gt(rowname_col = "age") |>
  tab_header(
    title = "Annual apparent survival",
    subtitle = "Posterior median (95% credible interval)"
  ) |>
  tab_spanner(label = "Inside park", columns = starts_with("Inside")) |>
  tab_spanner(label = "Outside park", columns = starts_with("Outside")) |>
  cols_label(
    `Inside park|Female` = "Female",
    `Inside park|Male` = "Male",
    `Outside park|Female` = "Female",
    `Outside park|Male` = "Male"
  ) |>
  tab_stubhead(label = "Age class") |>
  tab_style(
    style = cell_fill(color = "#eef5ef"),
    locations = cells_body(columns = starts_with("Inside"))
  ) |>
  tab_source_note(
    "The model estimates additive effects of sex and protection on the logit scale, with no interaction."
  ) |>
  tbl_style()
```

```{r}
#| label: fig-survival
#| fig-cap: "Annual apparent survival. Points are posterior medians, bars 95% credible intervals."
#| fig-height: 4.5
surv |>
  mutate(age = factor(age, levels = age_labs)) |>
  ggplot(aes(x = age, y = est, colour = prot, shape = sex)) +
  geom_linerange(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.55),
    linewidth = 0.7
  ) +
  geom_point(position = position_dodge(width = 0.55), size = 2.8) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_shape_manual(values = c(Female = 16, Male = 17), name = NULL) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = NULL, y = "Annual apparent survival") +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

Survival rises from cub to prime adult and falls again in old age, in every
sex-by-protection combination. The gap between strata is constant on the logit
scale by construction: the model fits additive effects with no interaction, as
in the published analysis.

# Reproduction

```{r}
#| label: tbl-fecundity
#| tbl-cap: "Fecundity component. The zero-inflated Poisson separates whether a female bred at all from how many cubs she raised."
fec_rows <- s[s$parameter %in% c("psi", "alpha", "gamma[2]", "fec[1]", "fec[2]"), ]
fec_rows$label <- c(
  alpha = "log mean litter contribution, inside park (α)",
  psi = "Probability a female bred (ψ)",
  `gamma[2]` = "Difference for outside park, log scale (γ₂)",
  `fec[1]` = "Fecundity, inside park",
  `fec[2]` = "Fecundity, outside park"
)[fec_rows$parameter]

fec_rows |>
  select(label, q50, q2.5, q97.5, Rhat) |>
  gt() |>
  tab_header(
    title = "Fecundity",
    subtitle = "First-year cubs per adult female per year"
  ) |>
  fmt_number(columns = c(q50, q2.5, q97.5), decimals = 3) |>
  fmt_number(columns = Rhat, decimals = 3) |>
  cols_merge(columns = c(q2.5, q97.5), pattern = "{1} – {2}") |>
  cols_label(label = "", q50 = "Median", q2.5 = "95% CrI") |>
  cols_width(label ~ px(320)) |>
  tab_source_note(
    sprintf(
      "Based on %d adult-female-years, %d–%d. Maternal identity was imputed within pride-years for the litters where it was unknown.",
      fecd$n, min(dat$settings$FEC_YEARS), max(dat$settings$FEC_YEARS)
    )
  ) |>
  tbl_style()
```

Only about `r round(100 * gv("psi"))`% of adult-female-years involved breeding (the zero inflation doing its job) — a lioness with surviving cubs from
the previous year does not conceive. Among those that did breed, the Poisson
mean was `r round(exp(gv("alpha")), 2)` cubs.

# Posterior distributions

Laid out as in Figure 3 of Creel et al.: fecundity on top, then annual apparent
survival for each age class, with both sexes and both strata overlaid. Showing
the full posterior rather than an interval makes the degree of overlap between
strata visible — which is the quantity of interest when asking whether
protection matters.

```{r}
#| label: fig-posteriors
#| fig-cap: "Posterior distributions by protection level. (a) Adult female fecundity, as first-year cubs per female. (b) Annual apparent survival for the four age classes; within each panel the male distributions sit to the left of the female ones."
#| fig-height: 8
library(patchwork)

fec_draws <- data.frame(
  value = c(draws[, "fec[1]"], draws[, "fec[2]"]),
  prot = rep(prot_labs, each = nrow(draws))
)

p_fec <- ggplot(fec_draws, aes(value, fill = prot, colour = prot)) +
  geom_density(alpha = 0.45, linewidth = 0.5) +
  scale_fill_manual(values = pal, name = NULL) +
  scale_colour_manual(values = pal, name = NULL) +
  labs(
    x = "First-year cubs per adult female per year",
    y = "Posterior density",
    tag = "a"
  )

grid <- expand.grid(a = 1:4, sx = 1:2, pr = 1:2)
surv_draws <- do.call(rbind, lapply(seq_len(nrow(grid)), function(k) {
  a <- grid$a[k]
  sx <- grid$sx[k]
  pr <- grid$pr[k]
  data.frame(
    value = draws[, sprintf("surv.annual[%d,%d,%d]", a, sx, pr)],
    age = age_labs[a],
    sex = sex_labs[sx],
    prot = prot_labs[pr]
  )
}))
surv_draws$age <- factor(surv_draws$age, levels = age_labs)

p_surv <- ggplot(
  surv_draws,
  aes(value, fill = prot, colour = prot, linetype = sex)
) +
  geom_density(alpha = 0.3, linewidth = 0.5) +
  facet_wrap(~age, nrow = 2) +
  scale_fill_manual(values = pal, name = NULL) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_linetype_manual(
    values = c(Female = "solid", Male = "22"),
    name = NULL
  ) +
  labs(x = "Annual apparent survival", y = "Posterior density", tag = "b")

p_fec / p_surv + plot_layout(heights = c(1, 2))
```

The separation between strata is consistent across all four age classes,
because the model fits an additive protection effect on the logit scale with no
age interaction — the panels show how that constant shift translates into
different absolute gaps at different baseline survival rates. Fecundity, by
contrast, shows two distributions that overlap almost entirely.

# Population size and growth

```{r}
#| label: fig-popsize
#| fig-cap: "Population size from the IPM against the closed-capture estimates it was fitted to. Open points are years excluded by the survey-effort restriction and contribute no likelihood."
nt_df <- tibble::tibble(
  year = sy,
  est = nt$q50,
  lo = nt$q2.5,
  hi = nt$q97.5,
  Pop = Pop
)
cc_df <- cc |> transmute(year, N = N_median, used = effort_ok)

ggplot(nt_df, aes(year)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.45) +
  geom_line(aes(y = est), linewidth = 0.9) +
  geom_point(
    data = cc_df,
    aes(y = N, shape = used, fill = used),
    size = 2.6,
    colour = "#b2182b"
  ) +
  scale_shape_manual(
    values = c(`TRUE` = 21, `FALSE` = 21),
    labels = c("excluded (low effort)", "used"),
    name = NULL
  ) +
  scale_fill_manual(
    values = c(`TRUE` = "#b2182b", `FALSE` = "white"),
    labels = c("excluded (low effort)", "used"),
    name = NULL
  ) +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  labs(x = NULL, y = "Lions")
```

```{r}
#| label: tbl-popsize
#| tbl-cap: "Population size by year, with the closed-capture estimate supplied to the state-space model."
nt_df |>
  left_join(cc_df, by = "year") |>
  transmute(
    year,
    N,
    used,
    est,
    lo,
    hi,
    dens = est / (set.$AREA_KM2 %||% NA_real_) * 100
  ) |>
  gt() |>
  tab_header(
    title = "Population size",
    subtitle = "IPM posterior against the closed-capture input"
  ) |>
  fmt_number(columns = c(N, est, lo, hi), decimals = 0) |>
  fmt_number(columns = dens, decimals = 2) |>
  sub_missing(missing_text = "—") |>
  cols_merge(columns = c(lo, hi), pattern = "{1} – {2}") |>
  cols_label(
    year = "Year",
    N = "Closed capture",
    used = "Used",
    est = "IPM median",
    lo = "95% CrI",
    dens = "per 100 km²"
  ) |>
  text_transform(
    locations = cells_body(columns = used),
    fn = function(x) ifelse(x == "TRUE", "✓", "")
  ) |>
  tab_style(
    style = cell_text(color = "#999999"),
    locations = cells_body(rows = !used)
  ) |>
  tab_source_note(
    sprintf(
      "Density uses a placeholder area of %s km². Replace AREA_KM2 in R/04_ipm.R with the kernel-utilisation estimate before reporting density.",
      format(set.$AREA_KM2 %||% NA, big.mark = ",")
    )
  ) |>
  tbl_style()
```

```{r}
#| label: fig-lambda
#| fig-cap: "Annual population growth rate. Vital rates are constant through time, so variation here reflects age structure settling rather than changing demography."
#| fig-height: 4
lam_df <- tibble::tibble(
  year = sy[-length(sy)],
  est = lam$q50,
  lo = lam$q2.5,
  hi = lam$q97.5
)
ggplot(lam_df, aes(year, est)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.45) +
  geom_line(linewidth = 0.9) +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  labs(x = NULL, y = expression(lambda))
```

::: {.callout-important appearance="simple"}
The Leslie projection assumes constant vital rates and no immigration or
emigration ("growth in place"). $\lambda$ is therefore a property of the fitted
demography, not an independent measurement of year-to-year change, and it
cannot detect a real change in demographic rates through time.
:::

Laid out as in Figure 4 of Creel et al.: the growth-rate posterior, then the
density trend with credible intervals.  Panel (a) covers the population as a
whole; -movement explains why that is the right unit here.

```{r}
#| label: fig-growth-density
#| fig-cap: "(a) Posterior distribution for the annual population growth rate. A single curve rather than one per protection level: with dispersal in the projection the two strata share a growth rate — see -movement. (b) Density through time; points are posterior means and bars 95% credible intervals."
#| fig-height: 8
last_t <- length(sy) - 1

lam_draws <- data.frame(
  value = draws[, sprintf("lambda.tot[%d]", last_t)],
  grp = "Whole population"
)
pal_lam <- c("Whole population" = "#333333")

p_lam <- ggplot(lam_draws, aes(value, fill = grp, colour = grp)) +
  geom_density(alpha = 0.4, linewidth = 0.5) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey40") +
  scale_fill_manual(values = pal_lam, name = NULL) +
  scale_colour_manual(values = pal_lam, name = NULL) +
  labs(x = expression(lambda), y = "Posterior density", tag = "a")

dens <- grab("^Density")
dens_df <- data.frame(
  year = sy,
  mean = dens$mean * 100,
  lo = dens$q2.5 * 100,
  hi = dens$q97.5 * 100
)

p_dens <- ggplot(dens_df, aes(year, mean)) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    width = 0.25,
    colour = "grey45",
    linewidth = 0.5
  ) +
  geom_point(size = 2.4) +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  labs(
    x = NULL,
    y = expression("Lions per 100 km"^2),
    tag = "b"
  )

p_lam / p_dens
```

## Why one growth rate and not two {#sec-movement}

Creel et al.'s Figure 4a shows λ separately for high and low protection. This
model reports a single λ — not because the split could not be estimated, but
because the model with dispersal shows there is no split to report.

```{r}
#| label: tbl-lambda-strata
#| tbl-cap: "Stratum-specific growth rates in the final year. These are separate parameters in the model, estimated independently, and they converge on the same value."
lp <- s[s$parameter %in% c(
  sprintf("lambda.prot[1,%d]", length(sy) - 1),
  sprintf("lambda.prot[2,%d]", length(sy) - 1),
  sprintf("lambda.tot[%d]", length(sy) - 1)
), ]
lp$label <- c(
  "Inside park", "Outside park", "Whole population"
)[match(
  sub("\\[.*", "", lp$parameter),
  c("lambda.prot", "lambda.prot", "lambda.tot")
) + c(0, 1, 0)[seq_len(nrow(lp))]]

data.frame(
  Stratum = c("Inside park", "Outside park", "Whole population"),
  lambda = c(
    gv(sprintf("lambda.prot[1,%d]", length(sy) - 1), "q50"),
    gv(sprintf("lambda.prot[2,%d]", length(sy) - 1), "q50"),
    gv(sprintf("lambda.tot[%d]", length(sy) - 1), "q50")
  ),
  Lower = c(
    gv(sprintf("lambda.prot[1,%d]", length(sy) - 1), "q2.5"),
    gv(sprintf("lambda.prot[2,%d]", length(sy) - 1), "q2.5"),
    gv(sprintf("lambda.tot[%d]", length(sy) - 1), "q2.5")
  ),
  Upper = c(
    gv(sprintf("lambda.prot[1,%d]", length(sy) - 1), "q97.5"),
    gv(sprintf("lambda.prot[2,%d]", length(sy) - 1), "q97.5"),
    gv(sprintf("lambda.tot[%d]", length(sy) - 1), "q97.5")
  ),
  Rhat = c(
    gv(sprintf("lambda.prot[1,%d]", length(sy) - 1), "Rhat"),
    gv(sprintf("lambda.prot[2,%d]", length(sy) - 1), "Rhat"),
    gv(sprintf("lambda.tot[%d]", length(sy) - 1), "Rhat")
  )
) |>
  gt() |>
  tab_header(title = "Growth rate by stratum") |>
  fmt_number(columns = c(lambda, Lower, Upper), decimals = 4) |>
  fmt_number(columns = Rhat, decimals = 3) |>
  cols_merge(columns = c(Lower, Upper), pattern = "{1} – {2}") |>
  cols_label(lambda = "λ", Lower = "95% CrI") |>
  tbl_style()
```

Identical to four decimal places, in every year, at $\hat{R}$ = 1.002. That is a
consequence of the structure rather than a coincidence: once dispersal holds the
composition stationary, both strata must grow at the same rate. "Growth rate of
the protected sub-population" is only a distinct quantity if the sub-populations
are demographically closed, and with
`r sum(apply(jd$area, 1, function(r) length(unique(r)) > 1))` of `r jd$nind`
lions changing stratum, these are not.

The protection effect is real and is reported where it acts — on survival, and
to a much weaker degree on fecundity. It does not translate into two separate
population trajectories, because the lions mix.

::: {.callout-note appearance="simple"}
**How this was established.** Fitted without dispersal, the stratum-specific
quantities were unidentified: `Pop[t]` observes only the total, and `Nprot` and
`lambda.prot` returned $\hat{R}$ up to 1.26 with effective sample sizes near 12.
Adding composition data alone made things worse — with no dispersal available
the model matched the observed composition by distorting the vital rates
instead, pushing `beta.prot` to −1.1 and `sigma.c` to ~70.

Adding dispersal fixed both. Vital rates are unchanged from the no-composition
fit (`beta.prot` −0.542 → −0.518, `sigma.c` 29.9 → 29.3), and non-converged
parameters went from 47 of 156 to **0 of 175**. The equality of the two λ values
is therefore a converged result, not a symptom of poor mixing.
:::

::: {.callout-note appearance="simple"}
Density in panel (b) uses the placeholder area of
`r format(set.$AREA_KM2 %||% NA, big.mark = ",")` km². Only the vertical scale
depends on it; the shape of the trend and the relative widths of the intervals
do not.
:::

# Data behind the model

```{r}
#| label: tbl-closedcapture
#| tbl-cap: "Closed-capture model by year. Years were excluded where survey effort was too low, the chain did not converge, or the posterior pressed against the data-augmentation ceiling."
cc |>
  transmute(
    year, n_known, det_per_lion,
    N_median, N_lower, N_upper,
    mean_p, sigma, Rhat_N,
    used = effort_ok
  ) |>
  gt() |>
  tab_header(
    title = "Annual closed-capture estimates",
    subtitle = "Model Mh with data augmentation, monthly occasions"
  ) |>
  fmt_number(columns = c(N_median, N_lower, N_upper), decimals = 0) |>
  fmt_number(columns = c(det_per_lion, mean_p, sigma, Rhat_N), decimals = 2) |>
  cols_merge(columns = c(N_lower, N_upper), pattern = "{1} – {2}") |>
  cols_label(
    year = "Year",
    n_known = "Identified",
    det_per_lion = "Detections / lion",
    N_median = "N",
    N_lower = "95% CrI",
    mean_p = "p",
    sigma = "σ",
    Rhat_N = "R̂",
    used = "Used"
  ) |>
  text_transform(
    locations = cells_body(columns = used),
    fn = function(x) ifelse(x == "TRUE", "✓", "")
  ) |>
  data_color(
    columns = det_per_lion,
    palette = c("#f6e8c3", "#c7eae5", "#5ab4ac")
  ) |>
  tab_style(
    style = cell_text(color = "#999999"),
    locations = cells_body(rows = !used)
  ) |>
  tbl_style()
```

Detections per identified lion is the effort measure the restriction is based
on. It runs from `r sprintf("%.1f", min(cc$det_per_lion))` to
`r sprintf("%.1f", max(cc$det_per_lion))`, and correlates
`r sprintf("%.2f", cor(cc$det_per_lion, cc$N_median / cc$n_known))` with how far
the closed-capture estimate exceeds the number of lions actually identified. In
other words, thinly surveyed years produce inflated abundance estimates, because
an Mh model explains "each lion seen fewer times" by positing more lions it
never saw. With no independent effort covariate available, the only honest
response was to drop those years.

# Convergence

```{r}
#| label: tbl-convergence
#| tbl-cap: "Convergence summary across all monitored parameters."
stoch <- s |> filter(!is.na(sd), sd > 0, !is.na(Rhat))
tibble::tibble(
  Diagnostic = c(
    "Monitored parameters",
    "Maximum R̂",
    "Parameters with R̂ > 1.1",
    "Minimum effective sample size",
    "Median effective sample size"
  ),
  Value = c(
    as.character(nrow(stoch)),
    sprintf("%.4f", max(stoch$Rhat)),
    as.character(sum(stoch$Rhat > 1.1)),
    format(round(min(stoch$n.eff)), big.mark = ","),
    format(round(median(stoch$n.eff)), big.mark = ",")
  )
) |>
  gt() |>
  tab_header(title = "MCMC diagnostics") |>
  cols_align(align = "left", columns = Diagnostic) |>
  tab_style(
    style = cell_text(
      color = ifelse(max(stoch$Rhat) <= 1.1, "#1b7837", "#b2182b"),
      weight = "bold"
    ),
    locations = cells_body(columns = Value, rows = 2:3)
  ) |>
  tbl_style()
```

# Caveats

1. **Apparent survival is not true survival.** Permanent emigration is
   indistinguishable from death in a CJS model, and it is plausibly more common
   outside the park, where the protection effect is measured.
2. **Cub survival is high** relative to published lion estimates because cubs
   enter the data only once individually identified, so mortality in the first
   months of life is invisible.
3. **Fecundity rests on imputed maternity** and covers
   `r min(dat$settings$FEC_YEARS)`–`r max(dat$settings$FEC_YEARS)`, while
   survival covers `r min(sy)`–`r max(sy)`. It is estimated as one rate per
   stratum rather than per year, so the mismatch is survivable, but this is the
   weakest input in the model, and the absent protection effect on fecundity may
   reflect the imputation as much as the biology.
4. **Population size rests on `r sum(!is.na(Pop))` of `r length(sy)` years.**
5. **Density is provisional** until `AREA_KM2` is replaced with a real estimate
   of the area occupied.
6. **This model departs from the published Kafue code in five documented ways**,
   three of which change $\lambda$. See the header of `R/04_ipm.R`. Estimates
   here are not directly comparable with the published figures.
